---
title: "corrplot vignette explore"
author: "p lovehagen"
date: "2023-02-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
```

## Intro

<span style="color: red;">This is based on fa_tuto.rmd</span>

First you will need to load, and potentially install the new packages that we are using today: `psych` and `GPArotation` and `paran`. The following code load the 5 libraries needed today and install them if needed (i.e. only if they are not already installed on the computer)

```{r, warning=FALSE, message=FALSE}
if(!require(psych)){
    install.packages("psych", repos = "https://www.stats.bris.ac.uk/R/")
    library(psych)
}

if(!require(GPArotation)){
    install.packages("GPArotation", repos = "https://www.stats.bris.ac.uk/R/")
    library(GOArotation)
}

if(!require(corrplot)){
    install.packages("corrplot", repos = "https://www.stats.bris.ac.uk/R/")
    library(corrplot)
}

if(!require(ade4)){
    install.packages("ade4", repos = "https://www.stats.bris.ac.uk/R/")
    library(ade4)
}

if(!require(factoextra)){
    install.packages("factoextra", repos = "https://www.stats.bris.ac.uk/R/")
    library(factoextra)
}

if(!require(paran)){
    install.packages("paran", repos = "https://www.stats.bris.ac.uk/R/")
    library(paran)
}
```

## Birds

We are going to use a dataset that we already saw in the PCA

```{r}
databird<-read.table("../data/birds.txt", h=T,sep="\t") # read tab separated table with headers
colnames(databird)<-c("total", "alar", "beakandhead", "humerus", "keel") #add colnames
```

1.  Using `summary` describe the dataset

```{r description, echo=FALSE}
# type your code here
summary(databird) # basic summary stats for each variable
which(is.na(rowSums(databird))) # which rows (if any) have missing data? by summing rows of is.na.  which returns Trues (i.e. where = 1)
which(!complete.cases(databird)) # returns logical vector where case is complete

databird<-databird[complete.cases(databird),] # only use complete cases

```

2.  The `psych` package offer a new function called `describe`, what does it tell you?

```{r}
#  look for help using ?describe
describe(databird,skew=FALSE)
```
the sd, se suggest that the birds are similar
49 observations



3.  Plot the correlation matrix using `corrplot` as in the previous tutorial.

```{r}
## type your code here
corrplot(cor(databird))
prettybirds<-databird
colnames(prettybirds) <- c("Total", "Alar", "Beak and Head", "Humerus", "Keel")
b <-cor(prettybirds)
#improvements - change text colour, labels, table col names
corrplot(b)
corrplot(b, order = 'AOE') # after 'AOE' reorder
corrplot(b, method = 'number') # colorful number
corrplot(b, method = 'color', order = 'alphabet')
corrplot(b, method = 'shade', order = 'AOE', diag = FALSE)
corrplot(b, method = 'square', order = 'FPC', type = 'lower', diag = FALSE)
corrplot(b, method = 'ellipse', order = 'AOE', type = 'upper')
corrplot.mixed(b, order = 'AOE')
corrplot.mixed(b, lower = 'shade', upper = 'pie', order = 'hclust', tl.col = 'black')
```

```{r}
prettybirds<-databird
colnames(prettybirds) <- c("Total", "Alar", "Beak and Head", "Humerus", "Keel")
corrplot(cor(databird), tl.col = "black", type = "lower")

# for options have a look at vignettes at cran corrplot