---
title: "R_FactorAnalysis_function"
author: "p lovehagen"
date: "2023-03-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r packages, include = FALSE}
# psych is used for factor analysis
if(!require(psych)){
    install.packages("psych", repos = "https://www.stats.bris.ac.uk/R/")
    library(psych)
}
# GPArotation is Gradient Projection Algorithm Rotation for FA
if(!require(GPArotation)){
    install.packages("GPArotation", repos = "https://www.stats.bris.ac.uk/R/")
    library(GOArotation)
}
# corrplot visualises correlation matrices
if(!require(corrplot)){
    install.packages("corrplot", repos = "https://www.stats.bris.ac.uk/R/")
    library(corrplot)
}
# ade4 is used for visualising and analysing pca
if(!require(ade4)){
    install.packages("ade4", repos = "https://www.stats.bris.ac.uk/R/")
    library(ade4)
}
# factoextra extracts and visualises results of multivariate data analyses
if(!require(factoextra)){
    install.packages("factoextra", repos = "https://www.stats.bris.ac.uk/R/")
    library(factoextra)
}
# paran applies Horn's Test of Principal Components / Factors
if(!require(paran)){
    install.packages("paran", repos = "https://www.stats.bris.ac.uk/R/")
    library(paran)
}
# ggplot2 is used to visualise

if(!require("ggplot2")){
  install.packages("ggplot2", repos = "https://www.stats.bris.ac.uk/R/")
  library(ggplot2)
}
```

## TODO

* add print statements to explain output
* add line breaks to make it more legible
* workout how to extract variables after running function

## Purpose

This is a function to allow for repeated `factor analysis`

It requires that there are no missing values in the dataframe



## Function inputs

* data = dataset
* num_factors = 3 is default
* rotation_method = "varimax" is default, options include "none", "varimax", "quartimax", "bentlerT", "equamax", "varimin", "geominT" and "bifactor" are orthogonal rotations. "Promax", "promax", "oblimin", "simplimax", "bentlerQ, "geominQ" and "biquartimin" and "cluster" 
* fm = "ml" is default (maximum likelihood factor analysis); options include "uls", "minres", "ols", "gls", "wls" - check help()
* scores = "regression" is default", options = "Thurstone", "tenBerge", "Anderson", "Bartlett" - see help()
* exclude_cols = NULL is default
* cutoff = 0.3 is default

## How it works

1. excludes columns in exclude_cols parameter, if provided
2. runs factor analysis (`fa()`) with parameters
3. prints loadings which are greater than `cutoff`
4. prints variance accounted for by each factor
5. prints communalities
6. plots biplot 

```{r}

# function to run fa
library(psych)

run_fa <- function(data, num_factors, rotation_method = "varimax", fm = "ml", scores = "regression", exclude_cols = NULL, cutoff = 0.3) {
  
  # remove columns specified in exclude_cols parameter
  if (!is.null(exclude_cols)) {
    data <- data[, !(colnames(data) %in% exclude_cols)]
  }
  
  # run factor analysis using specified method and rotation
  fa_obj <- fa(data, nfactors = num_factors, rotate = rotation_method, fm = fm, 
               scores = scores)
  
  # print factor loadings greater than or equal to the cutoff
  print(fa_obj$loadings[abs(fa_obj$loadings) >= cutoff])
  
  # print variance accounted for by each factor
  print(fa_obj$Vaccounted)
  
  # print communalities
  print(fa_obj$communalities)
  
  # create biplot of factor loadings and factor scores
  biplot.psych(fa_obj, main = paste("rotation:", rotation_method, ",", num_factors, "factors"))
}


```
## Example

```{r}

# data
USstates<- read.table("./_data/USstates.txt", header = TRUE) 
tmp <- USstates[, 2:9] 
rownames(tmp) <- USstates[, 1]
USstates <- tmp 
rm(tmp) 

# factor analysis
excluded_cols <- c("Population", "Area", "Frost")
fa_obj <- run_fa(USstates, 3, exclude_cols = excluded_cols)

```