---
title: "Package Managers in R - packrat, renv"
author: "P Lovehagen"
date: 2023-02-25
date-modified: "`r format(Sys.time(), '%Y-%m-%d')`"
tags: [R, process, package manager]
output:
  quarto::html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code:
      wrap: true
      eval: false
    css: "./ProcessStyle.css"
---

## **Packrat**

**`packrat`** is a package manager for R that allows you to create isolated, project-specific libraries of R packages. This can be useful for ensuring that your R code is reproducible across different environments and for managing dependencies for different projects. Here are the basic steps for using **`packrat`**:

### **Step 1: Initialize a `packrat` Project**

In your R project directory, run the following command to initialize a **`packrat`** project:

```{r}
codepackrat::init()
```

This will create a new directory called **`packrat`** in your project directory that will store all of the R packages that your project depends on.

### **Step 2: Install Packages**

To install R packages into your project-specific library, you can use the **`packrat::install()`** function. For example, to install the **`ggplot2`** package, you can run the following command:

```{r}
codepackrat::install("ggplot2")
```

This will download and install the **`ggplot2`** package into your project-specific library.

### **Step 3: Manage Dependencies**

To manage dependencies for your project, you can use the **`packrat::snapshot()`** function to create a snapshot of your project-specific library. This snapshot can be shared with others or used to recreate your project environment on a different machine. For example, to create a snapshot of your project library, you can run the following command:

```{r}
codepackrat::snapshot()
```

This will create a file called **`packrat.lock`** in your project directory that lists all of the R packages that your project depends on.

### **Step 4: Restore the Project Environment**

To restore your project environment on a different machine or after updating R, you can use the **`packrat::restore()`** function. This will install all of the R packages listed in the **`packrat.lock`** file into your project-specific library. For example:

```{r}
codepackrat::restore()
```

## **Renviron**

**`renv`** is another package manager for R that can help you manage your R package dependencies and ensure reproducibility across different environments. **`renv`** is similar to **`packrat`**, but it also includes additional features like support for Git-backed package sources and integration with continuous integration systems. Here are the basic steps for using **`renv`**:

### **Step 1: Initialize a `renv` Project**

In your R project directory, run the following command to initialize a **`renv`** project:

```{r}
coderenv::init()
```

This will create a new directory called **`.renv`** in your project directory that will store all of the R packages that your project depends on.

### **Step 2: Install Packages**

To install R packages into your project-specific library, you can use the **`renv::install()`** function. For example, to install the **`ggplot2`** package, you can run the following command:

```{r}
coderenv::install("ggplot2")
```

This will download and install the **`ggplot2`** package into your project-specific library.

### **Step 3: Manage Dependencies**

To manage dependencies for your project, you can use the **`renv::snapshot()`** function to create a snapshot of your project-specific library. This snapshot can be shared with others or used to recreate your project environment on a different machine. For example, to create a snapshot of your project library, you can run the following command:

```{r}
coderenv::snapshot()
```

This will create a file called **`renv.lock`**
