---
title: "Split datasets"
author: "p lovehagen"
date: "2023-04-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Purpose

This is a function to split a data set into three partitions - `train`, `test`, `validate` data sets.

It takes one argument `N` which is the number of indices, i.e. the length of the dataset.

There are three optional arguments to change the proportions with default values:

`train_prop` = 0.5

`test_prop` = 0.25

`val_prop` = 0.25

## Function

```{r}

# set a seed for reproducibility
set.seed(123)



partition_indices <- function(N, train_prop = 0.5, test_prop = 0.25, val_prop = 0.25) {
  
  # Create vector of indices
  Ind <- 1:N
  
  # Randomly select indices for training set
  train_size <- round(N * train_prop)
  trainInd <- sort(sample(Ind, train_size))
  
  # Create "rest" set of indices
  rest <- setdiff(Ind, trainInd)
  
  # Randomly select indices for testing set
  test_size <- round(N * test_prop)
  testInd <- sort(sample(rest, test_size))
  
  # Create validation set from remaining indices
  val_size <- length(rest) - test_size
  valInd <- rest[-match(testInd, rest)]
  
  # Return a list of indices for each set
  list(train = trainInd, test = testInd, validation = valInd)
}

```

```{r}
# function to partition datasets
partition_indices <- function(data, train_prop = 0.5, val_prop = 0.25, test_prop = 0.25, sort_by = "time") {
  
  # sort data by specified column/index
  sorted_data <- data[order(data[[sort_by]]),]
  
  # total number of rows in data
  N <- nrow(data)
  
  # vector of indices
  Ind <- 1:N
  
  # random select indices for training set
  train_size <- round(N * train_prop)
  trainInd <- sort(sample(Ind, train_size))
  
  # create "rest" set of indices
  rest <- setdiff(Ind, trainInd)
  
  # randomly select indices for testing set
  val_size <- round(N * val_prop)
  valInd <- sort(sample(rest, val_size))
  
  # create validation set from remaining indices
  test_size <- length(rest) - val_size
  testInd <- rest[-match(valInd, rest)]
  
  # Return a list of indices for each set
  list(train = trainInd,validation = valInd, test = testInd)
}
```

## to include stratification

```{r}
library(caret)

partition_indices <- function(data, target, train_prop = 0.5, val_prop = 0.25, test_prop = 0.25, stratified = FALSE) {
  
  # get total number of rows
  N <- nrow(data)
  
  # set seed for reproducibility
  set.seed(123)
  
  # create train, test, and validation sets
  if(stratified) {
    train_size <- round(N * train_prop)
    test_size <- round(N * test_prop)
    val_size <- N - train_size - test_size
    indices <- createDataPartition(data[[target]], p = c(train_prop, val_prop, test_prop), list = FALSE, times = 1)
    trainInd <- indices[1:train_size]
    valInd <- indices[(train_size+1):(train_size+val_size)]
    testInd <- indices[(train_size+val_size+1):(train_size+val_size+test_size)]
  } else {
    indices <- sample(1:N)
    train_size <- round(N * train_prop)
    val_size <- round(N * val_prop)
    trainInd <- indices[1:train_size]
    valInd <- indices[(train_size+1):(train_size+val_size)]
    testInd <- indices[(train_size+val_size+1):N]
  }
  
  # return a list of indices for each set
  list(train = trainInd, validation = valInd, test = testInd)
}

```


## How to use

```{r}

# Split 1000 indices into a 60/20/20 training/testing/validation split
partitioned_indices <- partition_indices(N = 1000, train_prop = 0.6, test_prop = 0.2, val_prop = 0.2)

# Access the indices for each set
train_indices <- partitioned_indices$train
test_indices <- partitioned_indices$test
val_indices <- partitioned_indices$validation

# Print the number of indices in each set
cat("Number of indices in training set:", length(train_indices), "\n")
cat("Number of indices in testing set:", length(test_indices), "\n")
cat("Number of indices in validation set:", length(val_indices), "\n")


```

## Example

Assume data frame called `sample_data`

```{r}

# Get number of rows in dataframe
N <- nrow(sample_data)

# call `partition_indices` function with N as the first argument
# changed props (if needed) to 0.7, 0.2, 0.1 

partitioned_indices <- partition_indices(N = nrow(sample_data), train_prop = 0.7, test_prop = 0.2, val_prop = 0.1)

# `particitioned_indices` contains indices for each of the three sets which is used to extract rows from `sample_data`
train_data <- sample_data[partitioned_indices$train, ]
test_data <- sample_data[partitioned_indices$test, ]
val_data <- sample_data[partitioned_indices$validation, ]


```

Result is three new dataframes which contain corresponding rows from `sample_data`
