---
title: "Upgrading R and RStudio"
author: "P Lovehagen"
date: 2023-02-25
date-modified: "`r format(Sys.time(), '%Y-%m-%d')`"
tags: [R, process guide, upgrade]
output:
  quarto::html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code:
      wrap: true
      eval: false
    css: "./ProcessStyle.css"
---

## 1. R Version

Check current R version and latest available version.

In RStudio type following code:

```{r}
R.Version()$version.string

```

This will display the current version of R.

To check the latest available version of R, go to CRAN website ([**https://cran.r-project.org/**](https://cran.r-project.org/)).

## 2. RStudio Version

Check current RStudio version and latest available version.

In RStudio, go to menu bar and select "Help" \> "About RStudio" which will display the version of RStudio in use, e.g. :

::: {#RStudio_version .special-block}
RStudio 2022.12.0+353 "Elsbeth Geranium" Release (7d165dcfc1b6d300eb247738db2c7076234f6ef0, 2022-12-03) for Windows

Mozilla/5.0 (Windows NT 10.0; Win64; x64) AppleWebKit/537.36 (KHTML, like Gecko) RStudio/2022.12.0+353 Chrome/102.0.5005.167 Electron/19.1.3 Safari/537.36
:::

To check the latest version of RStudio, go to the RStudio website ([**https://www.rstudio.com/products/rstudio/download/#download**](https://www.rstudio.com/products/rstudio/download/#download)) and check the version number. Or go to "Help" \> "Check for Updates"

## 3. Upgrading R {#sec-3.-upgrading-r}

Steps to upgrade R:

### Step 1: Download Latest Version of R

Go to the CRAN website (https://cran.r-project.org/) and download the latest version of R compatible with your operating system.

### Step 2: Install the New Version of R

Run the installer and follow the instructions to install the new version of R. During the installation process, you should choose a different directory for the new installation to avoid overwriting any files or configurations from the previous installation.

For example, to upgrade from `R-4.2.1` to `R-4.2.2` then the new default location may be `C:\Program Files\R\R-4.2.2`

### Step 3: Point RStudio to the New Version of R

After installing the new version of R, point RStudio to use the new version instead of the old one. To do this, open RStudio and go to the menu bar and select "Tools" \> "Global Options" \> "General". Under "R version", click on "Change" and browse to the new R installation directory. Select the R executable file (e.g., R.exe on Windows) and click "OK".

### Step 4: Migrate Libraries

When upgrading to a new version of R, you don't necessarily need to migrate packages. New version of R will be able to access the packages installed bythe previous version of R. However, some packages may need to be reinstalled to work properly

If migration is required, use `install.packages()` function with `lib.loc` argument to specify directory where packages are currently installed and the `repos` argument to specify the CRAN mirror to use.

```{r}
#To check path to the R installation directory
.libPaths()

#store locations
old_lib_loc <- "C:/path/to/old/r/library"
new_lib_loc <- "C:/path/to/new/r/library"

# repo locations, e.g. :
repos <- "https://cloud.r-project.org/"
#"https://www.stats.bris.ac.uk/R/"

installed_pkgs <- rownames(installed.packages(old_lib_loc))

install.packages(installed_pkgs, lib = new_lib_loc, repos = repos)


```

Or us can use a package manager such as `packrat` or `renv`. These package managers for R help manage R package dependencies and ensure reproducibility across different environments.

### Step 5: Remove Previous Version of R (Optional)

After verifying that the new version of R is working properly with RStudio, you can remove the previous version of R from your system. To do this, go to `Add / Remove Programs`

## 4. Upgrade RStudio

Steps to upgrade RStudio

### Step 1: Download Latest Version of RStudio

Go to the RStudio website (https://www.rstudio.com/products/rstudio/download/#download) and download the latest version of RStudio compatible with your operating system.

### Step 2: Install the New Version of RStudio

Run the installer and follow the instructions to install the new version of RStudio. During the installation process, you should choose the same directory as your current RStudio installation to ensure that your R projects and settings are still accessible.

### Step 3: Launch the New Version of RStudio

After installing the new version of RStudio, you can launch it from your desktop or command line. Your R projects and settings
